% -*- mode: fundamental -*-

% ****************************************************************

\chapter{Pending (to be written): Advanced topics, possibly in ``Book 2''?}

\markboth{Ch \arabic{chapter}: Pending (DRAFT)}{\copyrightnotice}

\setcounter{page}{1}
% \renewcommand{\thepage}{\arabic{page}}
\renewcommand{\thepage}{\arabic{chapter}-\arabic{page}}

\label{ch_Pending}

% ****************************************************************

This is a place-holder chapter with a TODO list of pending possible
topics to be written.

Goal: Linux capability, better performance.

\hdivider

% ****************************************************************

\section{Advanced branch prediction}

Is a form of online machine-learning.

Branch instruction hints.

Branch-Target buffers (BTBs)

Return-Address Stacks (RASs)

Hysteresis in prediction.

% ****************************************************************

\section{Caches}

Separate I- and D-caches.

FENCE.I: for ``manual'' I- and D-Mem coherence.

FENCE: flushing caches for for devices.

Multi-level caches and cache hierarchies.

Cache-coherence.

Non-blocking caches.

% ****************************************************************

\section{Compressed instructions}

Implementation: wholly in Decode stage.  Common shared between Fife and Drum.

Affect of compressed instructions on Fetch.

Affect of compressed instructions on PC-prediction.

Affect of non-32-bit alignment of compressed instructions.

% ****************************************************************

\section{AMO operations}

Note: These (implementation-independent) explanations may already be
done in Chapter~\ref{ch_ISA} on the RISC-V ISA:

\begin{itemize}
  \item Explanation of LR and SC instructions
  \item Explanation of AMOxxx instructions
  \item Implementation in the memory system
\end{itemize}

% ****************************************************************

\section{Multiple Privilege levels}

Machine, Supervisor, User

Interrupt/trap delegation.

Hypervisor

% ****************************************************************

\section{Memory Protection with PMPs}

Page Tables, TLBs

% ****************************************************************

\section{Virtual Memory}

Page Tables, TLBs

% ****************************************************************

\subsection{RISC-V ISA Formal Specification}

Sail model

% ****************************************************************
